---
title: "Predicting exercise effectiveness from personal activity data"
author: "John Fitzpatrick"
date: "Saturday, November 22, 2014"
output: html_document
---

## Executive Summary
#### The purpose of this analysis is to build a model from personal activity data taken from
#### devices such as Jawbone Up, Nike FuelBand, and Fitbit to help determine whether an
#### individual is performing a dumbbell biceps curl effectively. The effectiveness of the
#### model will be judged against a cross validation test and then finally applied to the
#### testing data which is a separately stored dataset provided by the authors
####
#### The data for this analysis comes from http://groupware.les.inf.puc-rio.br/har

**Download the training and test dataset**

```{r download_data,echo=TRUE}

dataDir <- "data"
trainFileName <- paste(dataDir,"pml-training.csv",sep="/")  
testFileName <- paste(dataDir,"pml-testing.csv",sep="/")  

if (!file.exists(dataDir)) {
  dir.create(dataDir)
  
  setInternet2(TRUE)  # set the R_WIN_INTERNET2 to TRUE  
  train.ds.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  download.file(train.ds.url,trainFileName)

  test.ds.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  download.file(test.ds.url,testFileName)
  
  dateDownloaded <- date()
  dateDownloaded
}


```

**Read in the training dataset**

```{r read_training,echo=TRUE}
training.ds <- read.csv(trainFileName,header=TRUE,stringsAsFactors=FALSE)

```

**Our goal is to build a model to predict on an unseen dataset the value of 
the "classe" variable**  

**This variable takes the following values**

* A - exercise performed exactly according to the specification  
* B - exercise performed throwing the elbows to the front  
* C - exercise performed  lifting the dumbbell only halfway  
* D - exercise performed lowering the dumbbell only halfway  
* E - exercise performed throwing the hips to the front  

**In looking through the dataset there are many columns that contain primarily NAs.**
**These won't contribute much to describing the data or helping us make predictions about 
unseen data. In addition, because we're trying to predict the class across all people and times,
variables relating to those parameters will be dropped**
```{r elim_poor_predictors,echo=TRUE}
training.ds$classe <- as.factor(training.ds$classe)
training.ds$X <- NULL
training.ds$user_name <- NULL
training.ds$new_window <- NULL
training.ds$num_window <- NULL
training.ds$raw_timestamp_part_1 <- NULL
training.ds$raw_timestamp_part_2 <- NULL
training.ds$cvtd_timestamp <- NULL

# convert the remaining character columns to numeric
# any values that can't be converted will be changed to NA
for (i in which(sapply(training.ds, is.character))) {
  training.ds[,i] <- as.numeric(training.ds[,i])    
}

# find columns that have greater than 40% NA
remove.threshold <- 0.5
total.rows <- nrow(training.ds)
remove.cols <- c()


for (i in 1:(ncol(training.ds) - 1)) {
  na.count <- sum(is.na(training.ds[[i]]))
  if ((na.count / total.rows) > remove.threshold) {
    remove.cols <- append(remove.cols,i)
  }
}

training.ds <- training.ds[,remove.cols * -1]


```

**Partition the training dataset so that we can cross validate our model once built**

```{r partition_data,echo=TRUE}
library(caret)
inTrain <- createDataPartition(training.ds$classe, p = 3/4)[[1]]
training <- training.ds[ inTrain,]
cv <- training.ds[-inTrain,]


```

**Build a model using the random forest algorithm. Because there are so many variables
preprocess the data using pca to reduce the number of factors needed to build the model**

```{r create_model,echo=TRUE}

set.seed(33833)
modelFit <- train(training$classe ~ .,method="rf",preProcess=c("center","scale","pca"),data=training)


```

**Assess the accuracy of the model by predicting the class using the cross validation data**

```{r cross_validate,echo=TRUE}
cv.prediction <- predict(modelFit,cv)
print(confusionMatrix(cv$classe,cv.prediction))
```

*The accuracy in the confusion matrix is 97% which means the out of sample error is 3%. The model should do a good job of predicting the class on the test dataset*


**Load the test dataset and perform the same preprocessing on it as was done on the 
training dataset**

```{r read_test,echo=TRUE}
testing.ds <- read.csv(testFileName,header=TRUE,stringsAsFactors=FALSE)

```

```{r elim_test_predictors,echo=TRUE}
testing.ds$X <- NULL
testing.ds$user_name <- NULL
testing.ds$new_window <- NULL
testing.ds$num_window <- NULL
testing.ds$raw_timestamp_part_1 <- NULL
testing.ds$raw_timestamp_part_2 <- NULL
testing.ds$cvtd_timestamp <- NULL
testing.ds$problem_id <- NULL

# convert the remaining character columns to numeric
# any values that can't be converted will be changed to NA
for (i in which(sapply(testing.ds, is.character))) {
  testing.ds[,i] <- as.numeric(testing.ds[,i])    
}

testing.ds <- testing.ds[,remove.cols * -1]


```

**Run the model against the test data to generate predictions**
```{r predict_test,echo=TRUE}
testing.prediction <- predict(modelFit,testing.ds)
print(testing.prediction)

```


## Conclusion
#### It is possible to predict the effectiveness of exercise based on personal activity data
#### with a high degree of accuracy.  

#### Based on the submission of these results against the actual values 19 out of 20 were
#### predicted correctly